Visualisation of Brain Statistics with R-package \texttt{ggseg}

### 2019-04-29 draft_2.tex
updated ms from comments and added Didacs discussion

\usepackage{lipsum}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

Neuroscientific analysis usually requires the use of multiple tools to analyse, visualise, and summarise data.
This often makes the process of preparing results for publication laborious, as results must be exported and imported in various formats.
In neuroimaging, images of probands' brains are collected and merged together to provide three dimensional representations.
Much of the neuroimaging analyses are not necessarily done on the individual voxel level (3-dimensional pixels), but rather on pre-defined brain segmentations, called brain atlases.
These brain atlases are plentiful and represent different ways of segmenting the brain into functionally or structurally similar regions (\citet{dkt}).
The use of these is wide-spread, as these atlases provide larger meaningful divisions of the brain.
While neuroimaging analyses on the voxel-level are usually computed by special software for such analyses, analyses of brain atlas data is usually done in standard statistical software, like R (\citet{R}), python (\citet{python}), or MATLAB (\citet{MATLAB}).

A key part of understanding and disseminating analysis results, is the visualization of these in a meaningful way.
With regards to results from brain atlas analyses, it is most meaningfully represented if projected onto a representation of the brain, rather than other common types of charts (like bar-charts) accompanied by a diagram of the brain label positions.
Each atlas has its own labelling depending on what is meaningful for the type of segmentation it is based on.
As such, for the reader to fully understand a bar chart with atlas labels, they need to be very familiar with the location of each label to have a clear comprehension of the results.
A projection directly onto a brain shape, eases the readability of the results for the reader, and provides clear point of reference even if the used atlas is unfamiliar.

There are several tools that aid R-users in plotting neuroimaging data directly through R using the grammar of graphics as implemented in ggplot2 (\citet{ggplot}), such as ggBrain (\citet{ggBrain}) and ggneuro (\citet{ggneuro}; see neuroconductor \citeyearpar{neuroconductor} for curated neuroimaging packages for R).
These are based on plotting imaging files, not results from analyses of brain atlases.
Here we introduce the ggseg-package for visualizing results from brain atlas analyses.
The ggseg-package was developed to create templates that others might use to project their brain atlas results on to.
While its plotting functions are what the users are drawn to, it is the pre-compiled number of data sets for different brain atlases that provides the real functionality needed for this visualization.
The package also includes functions to plot atlases in tri-surface mesh plots using plotly (\citet{plotly}).

\hypertarget{brain-atlas-selection}{%
\subsection{Brain atlas selection}\label{brain-atlas-selection}}

There is a multitude of brain atlases to chose from, all with different purposes and segmented based on different criteria.
We have initially focused on commonly used and well-established atlases with manageable numbers of segmentations.
The ggseg-package thus features data for two well-established parcellations, as implemented in the neuroimaging software Freesurfer \citep{fischl:99}; dkt and aseg.
The Desikan-Killany cortical atlas (dkt; \citet{dkt}) and the Automatic Segmentation of Subcortical Structures (aseg; \citet{aseg}) together cover all grey matter of the brain, and are commonly used in analyses elucidating the associations between thickness, volume, or area of these segments and mental health and cognitive function.

While functions and base data can be found in the main \texttt{ggseg}-package, a companion package called \texttt{ggsegExtra} has numerous other atlases, and is continuously expanded.
We encourage the community to help build and expand the atlases available, and to create a repository of established and often used atlases.
A summary of all available atlases may be found in Table \ref{tab:atlasTab}, where in addition to the two default atlases, ggsegExtra has another 12 atlases.
The \href{https://github.com/LCBC-UiO/ggseg/wiki}{ggseg wiki} currently has two different ways you may create and supply atlases to the package as 2D polygons.
Work is being done to include instructions on 3D-mesh plot contributions also.
The package includes functions like \texttt{as\_ggseg\_atlas()} and \texttt{as\_ggseg3d\_atlas()} to aid users in testing and making custom atlases.

@@

\caption{\label{tab:atlasTab}Table of currently available atlases in either ggseg or ggsegExtra R-package. Most atlases have both polygon and mesh atlases, but the mesh atlases are somewhat easier to create and are thus more plentiful. }
\centering
\begin{tabular}{lllll}
\toprule
Package & Title & Item & Mesh & Polygon\\
\midrule
 & Desikan-Killiany Cortical Atlas & dkt & No & Yes\\
\cmidrule{2-5}
 & \makecell[l]{Desikan-Killiany Cortical Atlas - tri-\\surf mesh Mesh data for the Desikan-\\Killany Cortical atlas, with 40 regions\\in on the cortical surface of the brain.} & dkt & Yes & No\\
\cmidrule{2-5}
\multirow{-3}{*}{\raggedright\arraybackslash ggseg} & \makecell[l]{Freesurfer automatic subcortical\\segmentation of a brain volume} & aseg & Yes & Yes\\
\cmidrule{1-5}
 & Desterieux cortical parcellations & desterieux & Yes & No\\
\cmidrule{2-5}
 & Harvard-Oxford Cortical atlas & hoCort & No & Yes\\
\cmidrule{2-5}
 & Parcellation from a midsagittal slice & midsagittal & No & Yes\\
\cmidrule{2-5}
 & Parcellation from JHU & jhu & No & Yes\\
\cmidrule{2-5}
 & Parcellation from of JHU & jhu & Yes & No\\
\cmidrule{2-5}
 & Parcellation from of white matter & icbm & Yes & No\\
\cmidrule{2-5}
 & Parcellation from of white matter & tracula & Yes & Yes\\
\cmidrule{2-5}
 & \makecell[l]{Parcellation from the Human Connectome\\Project} & glasser & Yes & Yes\\
\cmidrule{2-5}
 & \makecell[l]{Schaefer 17 Resting-state Cortical\\Parcellations} & schaefer17 & Yes & No\\
\cmidrule{2-5}
 & \makecell[l]{Schaefer 7 Resting-state Cortical\\Parcellations} & schaefer7 & Yes & No\\
\cmidrule{2-5}
 & \makecell[l]{Yeo 17 Resting-state Cortical\\Parcellations} & yeo17 & Yes & Yes\\
\cmidrule{2-5}
\multirow{-12}{*}{\raggedright\arraybackslash ggsegExtra} & \makecell[l]{Yeo 7 Resting-state Cortical\\Parcellations} & yeo7 & Yes & Yes\\
\bottomrule
\end{tabular}
\end{table}

@@

The aim of this tutorial is to familiarize the reader with the main package functions and the general use of the package.\\
It will cover the two main functions: \texttt{ggseg()} which plots 2D polygons, and \texttt{ggseg3d()} which plots tri-surface mesh plots.

\hypertarget{plotting-polygon-data}{%
\subsection{Plotting polygon data}\label{plotting-polygon-data}}

The function to plot two-dimensional polygon data is the \texttt{ggseg}-function.
This function will automatically plot the dkt-atlas by default (see Figure \ref{fig:init}).
As this function is a wrapper for \texttt{geom\_polygon} from \texttt{ggplot2} it can be built upon like all ggplot objects.
The plot is kept as simple as possible, with as little extra information to ggplot as possible.
This is done for it to be easy to extend the plot with the many ggplot options available.

@@

\begin{figure}
\centering
\includegraphics{draft_1_files/figure-latex/init-1.pdf}
\caption{\label{fig:init}By default \texttt{ggseg} will plot the dkt atlas in shaded polygons.}
\end{figure}

Because of the way \texttt{ggseg} is implemented, you should be able to safely use the function together with other ggplot functions, like themes, scales, and such (see Figure \ref{fig:void}).
There are many extra options to ggseg, to alter the plot as you wish.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{p1 <-}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{position =} \StringTok{"stacked"}\NormalTok{) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{theme_dark}\NormalTok{() }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{"dkt"}\NormalTok{, }\DataTypeTok{subtitle =} \StringTok{"dark theme"}\NormalTok{)}
\NormalTok{p2 <-}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{position =} \StringTok{"stacked"}\NormalTok{) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{theme_classic}\NormalTok{() }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{"dkt"}\NormalTok{, }\DataTypeTok{subtitle =} \StringTok{"classic theme"}\NormalTok{)}

\CommentTok{# Combine plots}
\NormalTok{pp <-}\StringTok{ }\NormalTok{cowplot}\OperatorTok{::}\KeywordTok{plot_grid}\NormalTok{(p1, p2, }\DataTypeTok{nrow =} \DecValTok{2}\NormalTok{)}

\NormalTok{p3 <-}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{atlas=}\NormalTok{aseg) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{"aseg"}\NormalTok{, }\DataTypeTok{subtitle =} \StringTok{"default theme"}\NormalTok{)}

\CommentTok{# Create a subplot grid}
\NormalTok{cowplot}\OperatorTok{::}\KeywordTok{plot_grid}\NormalTok{(pp, p3, }\DataTypeTok{nrow =} \DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{draft_1_files/figure-latex/void-1.pdf}
\caption{\label{fig:void}ggseg plots can be used with most standard scales, themes and such that work with ggplot. The special ggseg options for hemispheres, view etc. depend on the atlas used, and some options are only available for certain atlases. There is no `lateral' or `medial' views of subcortical atlases.}
\end{figure}

You can stack the hemispheres, view only the medial or lateral side, choose either or both of the hemispheres, or a combination of any of these (see Figure \ref{fig:stacked}).
The possibilities of altering the plot depends on the atlas used.
There are generally more options available for cortical atlases than subcortical ones, as the atlases have more meaningful ways of viewing the polygons.
Subcortical atlases also tend to have structures that span the hemispheres or other typically used referece points of the brain, and as such it is more difficult to label all segments as either, e.g., left or right.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{lat <-}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{view =} \StringTok{"medial"}\NormalTok{) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{"Medial view"}\NormalTok{)}
\NormalTok{left <-}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{hemisphere =} \StringTok{"left"}\NormalTok{) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{"Left hemi."}\NormalTok{)}
\NormalTok{pp <-}\StringTok{ }\NormalTok{cowplot}\OperatorTok{::}\KeywordTok{plot_grid}\NormalTok{(lat, left, }\DataTypeTok{nrow=}\DecValTok{2}\NormalTok{)}

\NormalTok{combo <-}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{view =} \StringTok{"medial"}\NormalTok{,}
      \DataTypeTok{hemisphere =} \StringTok{"left"}\NormalTok{) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{"Combination"}\NormalTok{)}

\NormalTok{cowplot}\OperatorTok{::}\KeywordTok{plot_grid}\NormalTok{(pp, combo, }\DataTypeTok{nrow=}\DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{draft_1_files/figure-latex/stacked-1.pdf}
\caption{\label{fig:stacked}There are several ggseg-special options that may be supplied to control how the plots looks and is organised. Top left: plot using medial view. Bottom left: plot using left hemisphere. Right: plot combining both the options.}
\end{figure}

@@

\texttt{ggseg} accepts any argument you can supply to \texttt{geom\_polygon} and therefore is easy to work with for those familiar with ggplot functionality.
Standard arguments like \texttt{fill} to flood the segments with a colour, or \texttt{colour} to colour the edges around the segments are typical arguments to provide to the function either as a single value or within the ggplot mapping option \texttt{aes}.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{mapping=}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{fill =}\NormalTok{ area), }\DataTypeTok{colour=}\StringTok{"black"}\NormalTok{) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{scale_fill_brain}\NormalTok{(}\StringTok{"dkt"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{legend.justification=}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{0}\NormalTok{),}
        \DataTypeTok{legend.position=}\StringTok{"bottom"}\NormalTok{,}
        \DataTypeTok{legend.text =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \DecValTok{5}\NormalTok{)) }\OperatorTok{+}
\StringTok{  }\KeywordTok{guides}\NormalTok{(}\DataTypeTok{fill =} \KeywordTok{guide_legend}\NormalTok{(}\DataTypeTok{ncol =} \DecValTok{3}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{draft_1_files/figure-latex/fill-1.pdf}
\caption{\label{fig:fill}Supplying `area' to the fill option in ggseg, will use the column `area' from the accompanying dataset to create a discrete colour palette over the segments in the atlas.}
\end{figure}

Typically you would want each area to have a different colour, corresponding to some descriptive or inference statistic that has been computed.
The bulk of the package originates not in its functions, but in the datasets that accompany the package.
There is one dataset per atlas, which contains key information regarding the atlas, and coordinates for the segment polygons of the atlas.
Before setting up the data from the statistics to project onto the segments, looking at the atlas datasets might help understand what the data needs to look like, or avoid looking like.

@@

The column `label' was added to the atlas for easy matching with the output from Freesurfer statistics tables.
The data is in long-format, meaning each segment has its own row, meaning the data you wish to project onto the segments also need to be in long-format.
As data directly from Freesurfer comes in wide-format, this requires some extra wranling.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{freesurfer_long <-}\StringTok{ }\NormalTok{freesurfer_stats }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{gather}\NormalTok{(label, thickness, }\OperatorTok{-}\NormalTok{id)}
\NormalTok{freesurfer_long}
\end{Highlighting}
\end{Shaded}

@@

This data can now be used directly with the \texttt{ggseg} function, as the \texttt{label} column corresponds in name and content with the \texttt{label} column in the atlas data of dkt.
The data \textbf{must} include a column that has the same name at atleast \emph{some} data matching the values in the corresponding column in the atlas data.
In the next example we create some data with 4 rows, and an `area' and `p' column.
The \texttt{ggseg} function will recognise the matching column `area', and merge the supplied data into the atlas using \texttt{dplyr} joins.
We use the \texttt{p} column as the column flooding the segment with colour.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{someData =}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}
  \DataTypeTok{area =} \KeywordTok{c}\NormalTok{(}\StringTok{"transverse temporal"}\NormalTok{, }\StringTok{"insula"}\NormalTok{,}
           \StringTok{"pre central"}\NormalTok{,}\StringTok{"superior parietal"}\NormalTok{), }
  \DataTypeTok{p =} \KeywordTok{sample}\NormalTok{(}\KeywordTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{,.}\DecValTok{001}\NormalTok{), }\DecValTok{4}\NormalTok{),}
  \DataTypeTok{stringsAsFactors =} \OtherTok{FALSE}\NormalTok{)}

@@

\includegraphics{draft_1_files/figure-latex/datasupp-1.pdf}

We can also change how this looks, by various ggplot functions, like scales, labs, themes and such.

@@

\includegraphics{draft_1_files/figure-latex/datasupp2-1.pdf}

If the results are only in one hemisphere, but you still want to plot both of them, make sure your data.fame includes the column \texttt{hemi} with either ``right'' or ``left'' for this to happen.

@@

\includegraphics{draft_1_files/figure-latex/datasuppX-1.pdf}

@@

If your data has different statistics for different groups, and you want to plot them all, we can use ggplot's \texttt{facet\_wrap} or \texttt{facet\_grid} for that.
The data needs to be in long format for this to work, as needed by ggplot, meaning the data from the groups should appear in seperate rows in the data, not in separate columns. Secondly, the data needs to be grouped using \texttt{dplyr}'s \texttt{group\_by} function \emph{before} providing the data to the \texttt{ggseg} function.
The ggseg-function will detect if the data is grouped, and perform the necessary steps for the \texttt{facet}'s to work when they are applied later.

>>>>

\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

Visualization is increasingly important for accurate guidance and interpretation of neuroimaging results, as current research is able to generate a high amount of data and outcomes.
For Magnetic Resonance Imaging (MRI), neuroimaging software provides whole-brain information by using many small units of space (\textgreater{}100.000).
Nonetheless, this data is often grouped and summarized into a limited number of regions using predefined brain parcellation atlases.
Brain parcellations segment the brain into a finite set of meaningful neurobiological components, which reflect one or more brain features either based on local/structural or connectivity properties (\citet{eickhoff_2018}).
The use of brain atlases is widespread as these facilitate interpretation and minimize the amount of data, hence reducing problems with multiple comparisons and facilitating replicability and data sharing in otherwise computationally expensive analyses.
Complex analyses and post-processing is often performed in specialized software environments such as R (\citet{R}).

MRI data provides good spatial resolution and thus an optimal representation has to respect spatial relationships across regions.
Results from brain atlas analyses are most meaningfully visualized when projected onto a representation of the brain, rather than on other types of graphs (e.g.~bar charts).
The projection of data into brain representations provides clear points of reference - especially when the reader is unfamiliar with the atlas - eases readability, guides interpretation, and informs on the spatial patterns of the data.
Adopting the grammar of graphics implemented in ggplot2 (\citet{ggplot}), one can plot neuroimaging data directly in R with several tools such as ggBrain (\citet{ggBrain}) and ggneuro (\citet{ggneuro}; see neuroconductor \citeyearpar{neuroconductor} for curated neuroimaging packages for R).
Yet these tools display whole-brain image files and are not suited for representing brain atlas data.

In this tutorial, we introduce the ggseg-package for visualizing brain atlas data in R.
The ggseg -- and the complimentary ggsegExtra -- package include pre-compiled data sets for different brain atlases that allow for 2D and 3D visualization.
Two-dimensional functionality is based on polygons and ggplot-based grammar of graphics (\citet{ggplot}), while the 3D functionality is based on tri-surface mesh plots and plotly (\citet{plotly}).

The ggseg-package presents both compiled data sets, tailored scripts to allow brain data integration and plotting and, other minor features such as custom color palettes.
The data featured are derived from two well-known parcellations: the Desikan-Killany cortical atlas (DKT; \citet{dkt}), which covers the cortical surface of the brain, and the Automatic Segmentation of Subcortical Structures (aseg; \citet{aseg}), which covers the subcortical structures.
Both atlases are implemented in several neuroimaging softwares, such as FreeSurfer (\citet{fischl_99}, \citet{dale_99}, \citet{Fischl2000}), and are commonly used in relation to developmental changes, disease biomarkers, genomic data, and cognition (\citet{amlien_elaboration_2019}, \citet{WALHOVD20051261}, \citet{Pizzagalli}).
The ggsegExtra-package currently includes \textgreater{}13 additional precompiled atlases and it is continuously expanding.
See Table \ref{tab:atlasTab} for a summary of the available atlases.

We encourage users to contribute to the ggsegExtra brain atlas repository.
In SI text and \href{https://github.com/LCBC-UiO/ggseg/wiki}{github wiki}, we offer a pipeline to create and supply atlases for 2D plotting.
See SI text and SI scripts for instruction to include atlases for 3D plotting.

@@

\caption{\label{tab:atlasTab}Table of currently available atlases in either ggseg or ggsegExtra R-package. Most atlases have both polygon and mesh atlases, but the mesh atlases are somewhat easier to create and are thus more plentiful.}
\centering
\begin{tabular}{lllll}
\toprule
Package & Title & Item & Mesh & Polygon\\
\midrule
 & Desikan-Killiany Cortical Atlas & dkt & Yes & Yes\\
\cmidrule{2-5}
\multirow{-2}{*}{\raggedright\arraybackslash ggseg} & \makecell[l]{Freesurfer automatic subcortical\\segmentation of a brain volume} & aseg & Yes & Yes\\
\cmidrule{1-5}
 & Desterieux cortical parcellations & desterieux & Yes & No\\
\cmidrule{2-5}
 & \makecell[l]{Genetic topography of brain area\\morphology} & chenAr & No & Yes\\
\cmidrule{2-5}
 & \makecell[l]{Genetic topography of brain thickness\\morphology} & chenTh & No & Yes\\
\cmidrule{2-5}
 & Harvard-Oxford Cortical atlas & hoCort & No & Yes\\
\cmidrule{2-5}
 & Parcellation from a midsagittal slice & midsagittal & No & Yes\\
\cmidrule{2-5}
 & Parcellation from JHU & jhu & Yes & Yes\\
\cmidrule{2-5}
 & Parcellation from of white matter & icbm & Yes & No\\
\cmidrule{2-5}
 & \makecell[l]{Parcellation from the Human Connectome\\Project} & glasser & Yes & Yes\\
\cmidrule{2-5}
 & \makecell[l]{Schaefer 17 Resting-state Cortical\\Parcellations} & schaefer17 & Yes & No\\
\cmidrule{2-5}
 & \makecell[l]{Schaefer 7 Resting-state Cortical\\Parcellations} & schaefer7 & Yes & No\\
\cmidrule{2-5}
 & White matter tract parcellations & tracula & Yes & Yes\\
\cmidrule{2-5}
 & \makecell[l]{Yeo 17 Resting-state Cortical\\Parcellations} & yeo17 & Yes & Yes\\
\cmidrule{2-5}
\multirow{-13}{*}{\raggedright\arraybackslash ggsegExtra} & \makecell[l]{Yeo 7 Resting-state Cortical\\Parcellations} & yeo7 & Yes & Yes\\
\bottomrule
\end{tabular}
\end{table}

@@

This tutorial will introduce the \texttt{ggseg} and \texttt{ggsegExtra} packages and familiarize the reader with the main functions and the general use of the packages.
The tutorial will cover the following functions: \texttt{ggseg()} for plotting 2D polygons and \texttt{ggseg3d()} for plotting 3D brains based on tri-surface mesh plots.

\hypertarget{plotting-polygon-data-ggplot2}{%
\subsection{Plotting polygon data (ggplot2)}\label{plotting-polygon-data-ggplot2}}

\texttt{ggseg} is the main function for plotting 2D data.
By default, the function automatically plots the DKT atlas (see Figure \ref{fig:init}).
The \texttt{ggseg} function is a wrapper for \texttt{geom\_polygon} from \texttt{ggplot2}, and it can be built upon and combined like any ggplot object.
The image plot consists of a simple brain representation containing no extra information.
Hence, \texttt{ggseg} plots can be easily complemented with any of the available ggplot features and options.
We recommend users to get familiarized with ggplot2 grammar (\citet{ggplot}).

@@

\begin{figure}
\centering
\includegraphics{draft_2_files/figure-latex/init-1.pdf}
\caption{\label{fig:init}By default \texttt{ggseg} will plot the dkt atlas in shaded polygons.}
\end{figure}

\texttt{ggseg} is implemented as a fully-functional part of ggplot.
Thus one should be able to use the function together with any other ggplot function such as themes or scales.
You can stack the hemispheres, view only the medial or lateral side, choose either or both of the hemispheres, or a combination of any of these.
One has several options for plotting the main brain representations.
These options are atlas-specific.
For cortical atlases, such as the `dkt', one can stack the hemispheres, plot one or both hemisphere, or view only the medial or lateral side (or any combination of these; see Figure \ref{fig:collection}).
For subcortical atlases, such as the `aseg', the options are more limited but one can often choose between axial, sagital, and coronal views.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# dkt dark theme}
\NormalTok{p1 <-}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{position =} \StringTok{"stacked"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{theme_dark}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{" "}\NormalTok{)}

\CommentTok{# dkt classic theme}
\NormalTok{p2 <-}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{position =} \StringTok{"stacked"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{theme_classic}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{" "}\NormalTok{)}

\CommentTok{# dkt medial view}
\NormalTok{med <-}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{view =} \StringTok{"medial"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{" "}\NormalTok{)}

\CommentTok{# dkt left hemisphere}
\NormalTok{left <-}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{hemisphere =} \StringTok{"left"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{" "}\NormalTok{)}

\CommentTok{# aseg default theme}
\NormalTok{p3 <-}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{atlas=}\NormalTok{aseg) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{" "}\NormalTok{)}

\CommentTok{# dkt left medial alone}
\NormalTok{combo <-}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{view =} \StringTok{"medial"}\NormalTok{,}
      \DataTypeTok{hemisphere =} \StringTok{"left"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{" "}\NormalTok{)}

\CommentTok{# Combine plots}
\NormalTok{cowplot}\OperatorTok{::}\KeywordTok{plot_grid}\NormalTok{(p1, med, combo,  p2, left, p3,}
                   \DataTypeTok{labels =} \KeywordTok{c}\NormalTok{(}\StringTok{"A: dkt - dark"}\NormalTok{, }\StringTok{"B: dkt - medial"}\NormalTok{,}
                              \StringTok{"C: dkt - combo"}\NormalTok{, }\StringTok{"D: dkt - classic"}\NormalTok{,}
                              \StringTok{"E: dkt - left"}\NormalTok{, }\StringTok{"F: aseg"}\NormalTok{),}
                   \DataTypeTok{hjust =} \FloatTok{-.05}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{draft_2_files/figure-latex/collection-1.pdf}
\caption{\label{fig:collection}ggseg plots can be used with most standard scales, themes and such that work with ggplot. The special ggseg options for hemispheres, view etc. depend on the atlas used, and some options are only available for certain atlases. There is no `lateral' or `medial' views of subcortical atlases. There are several ggseg-special options that may be supplied to control how the plots looks and is organised. \textbf{A:} dkt atlas, stacked with dark theme ; \textbf{B:} dkt with medial view only; \textbf{C:} dkt atlas with only left medial display; \textbf{D:} dkt atlas, stacked, with classic theme; \textbf{E:} dkt atlas with left hemisphere only; \textbf{F:} complete aseg atlas}
\end{figure}

@@

\texttt{ggseg} accepts any argument you can supply to \texttt{geom\_polygon} and therefore is easy to work with for those familiar with ggplot functionality.
Standard arguments like \texttt{fill} to flood the segments with a colour, or \texttt{colour} to colour the edges around the segments are typical arguments to provide to the function either as a single value or within the ggplot mapping option \texttt{aes}.
To use color palettes corresponding to those used in the original neuroimaging softwares one can use atlas-specific `brain' palette scales.
See an example in Figure \ref{fig:fill}.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{mapping=}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{fill =}\NormalTok{ area), }\DataTypeTok{colour=}\StringTok{"black"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale_fill_brain}\NormalTok{(}\StringTok{"dkt"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{legend.justification=}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{0}\NormalTok{),}
        \DataTypeTok{legend.position=}\StringTok{"bottom"}\NormalTok{,}
        \DataTypeTok{legend.text =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \DecValTok{5}\NormalTok{)) }\OperatorTok{+}
\StringTok{  }\KeywordTok{guides}\NormalTok{(}\DataTypeTok{fill =} \KeywordTok{guide_legend}\NormalTok{(}\DataTypeTok{ncol =} \DecValTok{3}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{draft_2_files/figure-latex/fill-1.pdf}
\caption{\label{fig:fill}Supplying `area' to the fill option in ggseg, will use the column `area' from the accompanying dataset to create a discrete colour palette over the segments in the atlas.}
\end{figure}

Most users will use ggseg to display - using a color scale - some descriptive or inferential statistics, such as mean thickness or cognition relationships etc., across the different brain regions.
The bulk of the package originates not in its functions, but in the data sets that accompany the package.
There is one data set per atlas, which contains key information regarding the atlas, and coordinates for the segment polygons of the atlas.
Before setting up the data from the statistics to project onto the segments, looking at the atlas data sets might help understand what the data needs to look like, or avoid looking like.

@@

In any atlas, the column `label' is particularly useful for combining the data of interest with the \texttt{ggseg}-polygons.
The column `label' contains the label (region) names as in the original neuroimaging software.
For example, the DKT atlas label column matches the region names from Freesurfer statistics tables.
Yet the data in \texttt{ggseg} is in a long format - that is each region has its own row - and data of interest needs to be in this same format.
Most data sets are organized in wide format, in which subjects are represented by rows and each different data variable is represented in a separate column, and thus need to be rearranged in order to work with ggseg.
See below an example of wide-to-long conversion.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{freesurfer_long <-}\StringTok{ }\NormalTok{freesurfer_stats }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{gather}\NormalTok{(label, thickness, }\OperatorTok{-}\NormalTok{id)}
\NormalTok{freesurfer_long}
\end{Highlighting}
\end{Shaded}

@@

This data - in long format - can now be used directly with the \texttt{ggseg} function, as the \texttt{label} column corresponds in name and content with the \texttt{label} column in the atlas data of dkt.
The data \textbf{must} include a column that has the same name and at least \emph{some} data matching the values in the corresponding column in the atlas data.
In the next example we create some data with 4 rows, and an `area' and `p' column, representing the results of an analysis.
The \texttt{ggseg} function will recognise the matching column `area', and merge the supplied data into the atlas using \texttt{dplyr} joins.
We use the \texttt{p} column as the column flooding the segment with colour.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{someData =}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}
  \DataTypeTok{area =} \KeywordTok{c}\NormalTok{(}\StringTok{"transverse temporal"}\NormalTok{, }\StringTok{"insula"}\NormalTok{,}
           \StringTok{"pre central"}\NormalTok{,}\StringTok{"superior parietal"}\NormalTok{),}
  \DataTypeTok{p =} \KeywordTok{sample}\NormalTok{(}\KeywordTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{,.}\DecValTok{001}\NormalTok{), }\DecValTok{4}\NormalTok{),}
  \DataTypeTok{stringsAsFactors =} \OtherTok{FALSE}\NormalTok{)}

@@

\includegraphics{draft_2_files/figure-latex/datasupp-1.pdf}

The appearance of this plot can then be modified similarly to any other ggplot graph using functions such as scales, labs, themes, etc.

@@

\includegraphics{draft_2_files/figure-latex/datasupp2-1.pdf}

If the results are only in one hemisphere, but you still want to plot both of them, make sure your data.fame includes the column \texttt{hemi} with either `right' or `left' for this to happen.

@@

\includegraphics{draft_2_files/figure-latex/datasuppX-1.pdf}

@@

There is often the need to plot a statistic of interest in different groups (i.e.~thickness or significance in young or older adults).
This may be obtained also with ggseg, using ggplot's \texttt{facet\_wrap} or \texttt{facet\_grid}, using two guiding rules:
\textbf{1)} as before, data needs to be in long format (group data should appear in seperate rows, not in separate columns).
\textbf{2)} The data needs to be grouped using \texttt{dplyr}'s \texttt{group\_by} function \emph{before} providing the data to the \texttt{ggseg} function.
The ggseg function will detect grouped data, and adapt it to \texttt{facet}'s requirements.
